Open-source tools in R for landscape ecology

### 2021-04-25 Comments_Reviewer.Rmd
:construction_worker: Addressing more comments...only a few left

@@

@@

@@

@@

@@

@@

>>>>

**Reply**: We agree that it was not clear enough how these package could help woth reproducibility, thus we added a more detailed explanation. \vskip 0.5cm

@@

**Reply**: Thank you for pointing out that any further explanation is missing. We added a brief explanation of what the tidyverse is and why using it can be beneficial. \vskip 0.5cm

@@

**Reply**: We added an explanation how these package allow to incorporate functionality currently only available in GIS software into reproducible and shareable R scripts \vskip 0.5cm

@@

**Reply**: @Jakub? \vskip 0.5cm

@@

**Reply**: We agree that this is problematic and we weren't even aware this is happeing so thanks for pointing it out. We deactivated the automatic hyphenation by LaTeX. \vskip 0.5cm

@@

**Reply**: Thank you for checking the table so extensively. We made sure all packages should be included now. \textcolor{red}{Check again all packages are there} \vskip 0.5cm

@@

**Reply**: We agree with the reviewer that there are many reasons to use other software or programming languages. Nevertheless, we think that the scope of this review is not to discuss pro and cons between the different tools, but rather to focus on R. At the same time, we believe that the manuscript does give many arguments why R might be a good solution for many tasks. \vskip 0.5cm

### 2021-04-25 paper/paper.Rmd
:construction_worker: Addressing more comments...only a few left

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g. Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The *R* programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig. \@ref(fig:fig-workflow)).
The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.

@@

The guaranteed technical quality standard of *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y*.
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
Furthermore, online communities like *rOpenSci* also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
For more information about *R* package development in general, see \cite{Wickham2015}.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 *R* packages directly depend or imports *sp*.
However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
*sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.

@@

Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}. 
\textcolor{blue}{Fore more information about **ggplot2** in general see \cite{Wickham2016}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
Both, *tmap* (Fig. \@ref(fig:fig-map)c) and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

@@

\caption{Overview of commonly used \textit{R} packages for spatial data and landscape ecology. Packages are sorted by their major application task.}

@@

>>>>

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g. Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The *R* programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig. \@ref(fig:fig-workflow)).
The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.

@@

The guaranteed technical quality standard of *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y*.
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
Furthermore, online communities like *rOpenSci* also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
For more information about *R* package development in general, see \cite{Wickham2015}.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 *R* packages directly depend or imports *sp*.
However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
\textcolor{blue}{The *tidyverse* is a collection of *R* packages developed for almost all major tasks of any data science project. 
Because all *tidyverse* packages follow the same data philosophy structures, one strength of the *tidyverse* is its high consistency of usage across its packages \cite{Wickham2019}.}
*sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.
\textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}

@@

Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}. 
\textcolor{blue}{Fore more information about *ggplot2* in general see \cite{Wickham2016}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
Both, *tmap* (Fig. \@ref(fig:fig-map)c) and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

@@

\captionsetup{width=2\linewidth}

\caption{Overview of commonly used \textit{R} packages for spatial data and landscape ecology. Packages are sorted by their major application task. The table only includes packages specifically desinged for spatial data or landscape ecology.}

@@

\multirow{3}{*}{Bridge stuff} & rgrass7 & Interface between GRASS7 and R  & \cite{Bivand2021} \\
& RSAGA & Interface between SAGA and R & \cite{Brenning2018} \\
& rgee &  Download data from Google Earth Engine & \cite{Aybar2020} \\

### 2021-04-25 paper/paper.tex
:construction_worker: Addressing more comments...only a few left

%% load any required packages here
\usepackage{caption}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{placeins}
\usepackage[modulo]{lineno}
\usepackage{xcolor}

@@

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The \emph{R} programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig. \ref{fig:fig-workflow}).
The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.

@@

The guaranteed technical quality standard of \emph{CRAN} requires more initial work for developers compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y}.
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use \emph{GitHub} for regular development and \emph{CRAN} to publish stable releases of the packages.
Furthermore, online communities like \emph{rOpenSci} also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include \emph{groundhog} \cite{Simonsohn2021}, \emph{packrat} \cite{Ushey2018}, or \emph{renv} \cite{Ushey2020}.
For more information about \emph{R} package development in general, see \cite{Wickham2015}.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The \emph{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \emph{R} packages directly depend or imports \emph{sp}.
However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019}.
\emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from \emph{R} using dedicated packages, such as \emph{rgrass7} \cite{Bivand2021} for GRASS GIS, \emph{RSAGA} \cite{Brenning2018} for SAGA GIS, and \emph{rgee} \cite{Aybar2020} for Google Earth Engine.

@@

Further plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
Also the popular plotting package \emph{ggplot2} \cite{Wickham2016} (Fig. \ref{fig:fig-map}b) has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{Dunnington2020}.
\textbackslash textcolor\{blue\}\{Fore more information about \textbf{ggplot2} in general see \cite{Wickham2016}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{Appelhans2020}.
Both, \emph{tmap} (Fig. \ref{fig:fig-map}c) and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{Cheng2021}.
A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

>>>>

%% load any required packages here
\usepackage{caption}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{placeins}
\usepackage[modulo]{lineno}
\usepackage{xcolor}
\usepackage[none]{hyphenat}

@@

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The \emph{R} programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig. \ref{fig:fig-workflow}).
The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.

@@

The guaranteed technical quality standard of \emph{CRAN} requires more initial work for developers compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y}.
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use \emph{GitHub} for regular development and \emph{CRAN} to publish stable releases of the packages.
Furthermore, online communities like \emph{rOpenSci} also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include \emph{groundhog} \cite{Simonsohn2021}, \emph{packrat} \cite{Ushey2018}, or \emph{renv} \cite{Ushey2020}.
\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
For more information about \emph{R} package development in general, see \cite{Wickham2015}.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The \emph{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \emph{R} packages directly depend or imports \emph{sp}.
However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019}.
\textcolor{blue}{The *tidyverse* is a collection of *R* packages developed for almost all major tasks of any data science project. 
Because all *tidyverse* packages follow the same data philosophy structures, one strength of the *tidyverse* is its high consistency of usage across its packages \cite{Wickham2019}.}
\emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from \emph{R} using dedicated packages, such as \emph{rgrass7} \cite{Bivand2021} for GRASS GIS, \emph{RSAGA} \cite{Brenning2018} for SAGA GIS, and \emph{rgee} \cite{Aybar2020} for Google Earth Engine.
\textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}

@@

Further plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
Also the popular plotting package \emph{ggplot2} \cite{Wickham2016} (Fig. \ref{fig:fig-map}b) has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{Dunnington2020}.
\textbackslash textcolor\{blue\}\{Fore more information about \emph{ggplot2} in general see \cite{Wickham2016}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{Appelhans2020}.
Both, \emph{tmap} (Fig. \ref{fig:fig-map}c) and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{Cheng2021}.
A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
